---
title: "Cross-Basis Projection Feature Encoding (CBPFE): A DLNM-Driven XGBoost Framework for Dengue Forecasting"
author: "Abrar Islam, Syed Shafkat Raiyan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

# Overview

This notebook implements the **Cross-Basis Projection Feature Encoding (CBPFE)** pipeline
described in Islam & Raiyan (2026). The pipeline has two stages:

- **Stage 1**: Four univariate NB-DLNMs extract nonlinear exposure-lag-response surfaces
  for mean temperature, precipitation, relative humidity, and wind speed.
- **Stage 2**: Cross-basis partial effects serve as engineered features for XGBoost
  forecasting, evaluated on a true 24-month prospective holdout (2024–2025).

All DLNM models used for feature extraction are refit on the training partition only
(2008–2023) to prevent data leakage into the test set.

---

# 1. Dependencies

```{r packages}
install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE))
    install.packages(pkg, repos = "https://cloud.r-project.org")
}

pkgs <- c("dlnm", "splines", "MASS", "readxl", "dplyr", "lubridate",
          "lmtest", "zoo", "car", "xgboost", "ggplot2", "tidyr",
          "forecast", "SHAPforxgboost", "moments", "scales")
invisible(lapply(pkgs, install_if_missing))

library(dlnm);        library(splines);    library(MASS)
library(readxl);      library(dplyr);      library(lubridate)
library(lmtest);      library(car);        library(zoo)
library(xgboost);     library(ggplot2);    library(tidyr)
library(forecast);    library(SHAPforxgboost)
library(moments);     library(scales)
```

---

# 2. Data

## 2.1 Load and Prepare

Replace the file paths below with your own data locations.
Expected columns after loading: `Date`, `DengueCases`, `MeanTemp`, `Precipitation`,
`Humidity`, `WindSpeed`. Population is merged from a separate annual CSV.

```{r load-data}
# --- Bangladesh dengue + climate dataset ---
df_raw <- read_excel("dataset/dengue_forecasting_dataset.xlsx", skip = 1)
colnames(df_raw) <- c("Date", "DengueCases", "MeanTemp", "Precipitation",
                      "Humidity", "WindSpeed", "DengueCases_plus1", "LN_DengueCases")
df_raw$Date      <- as.Date(df_raw$Date)
df_raw           <- df_raw %>% arrange(Date)
df_raw$Year      <- year(df_raw$Date)
df_raw$Month_num <- month(df_raw$Date)
df_raw <- df_raw %>%
  filter(!is.na(DengueCases), !is.na(MeanTemp),
         !is.na(Precipitation), !is.na(Humidity), !is.na(WindSpeed))

# --- Annual population estimates (World Bank) ---
pop_df <- read.csv("dataset/pop_dataset.csv", fileEncoding = "UTF-8-BOM")
colnames(pop_df)[1] <- "Year"
pop_df$Population   <- as.numeric(gsub(",", "", pop_df$Population))
df_raw <- df_raw %>% left_join(pop_df, by = "Year")
df_raw$Population <- zoo::na.locf(df_raw$Population, na.rm = FALSE)
df_raw$Population <- zoo::na.locf(df_raw$Population, fromLast = TRUE, na.rm = FALSE)
if (any(is.na(df_raw$Population))) stop("NA in Population after fill")

df_raw           <- as.data.frame(df_raw)
df_raw$Month_index <- seq_len(nrow(df_raw))

cat("Rows:", nrow(df_raw), "| Range:",
    format(min(df_raw$Date)), "to", format(max(df_raw$Date)), "\n")
```

## 2.2 Descriptive Statistics

```{r descriptive-stats}
vars_df <- df_raw[, c("DengueCases", "MeanTemp", "Precipitation",
                      "Humidity", "WindSpeed", "Population")]

desc_stats <- data.frame(
  Variable = c("Dengue cases (monthly count)", "Mean temperature (°C)",
               "Precipitation (mm)", "Relative humidity (%)",
               "Wind speed (m/s)", "Population (millions)"),
  Mean     = sapply(vars_df, function(x) round(mean(x, na.rm = TRUE), 2)),
  SD       = sapply(vars_df, function(x) round(sd(x, na.rm = TRUE), 2)),
  Median   = sapply(vars_df, function(x) round(median(x, na.rm = TRUE), 2)),
  IQR      = sapply(vars_df, function(x) round(IQR(x, na.rm = TRUE), 2)),
  Min      = sapply(vars_df, function(x) round(min(x, na.rm = TRUE), 2)),
  Max      = sapply(vars_df, function(x) round(max(x, na.rm = TRUE), 2)),
  Skewness = sapply(vars_df, function(x) round(skewness(x, na.rm = TRUE), 2)),
  Kurtosis = sapply(vars_df, function(x) round(kurtosis(x, na.rm = TRUE), 2))
)

# Convert population row to millions
pop_idx <- which(desc_stats$Variable == "Population (millions)")
desc_stats[pop_idx, c("Mean","SD","Median","IQR","Min","Max")] <-
  desc_stats[pop_idx, c("Mean","SD","Median","IQR","Min","Max")] / 1e6

knitr::kable(desc_stats, caption = "Table 1: Descriptive statistics (2008–2025, n=216 months)",
             digits = 2, align = "c")
```

## 2.3 Data Visualisation

### 2.3.1 Time-Series Overview

Monthly trajectories of all five study variables across the full 2008–2025 observation
window. The LOESS smoother (red dashed line) highlights the long-run trend within
each panel. The dengue panel uses a $\log_{1p}$ scale to prevent the 2023 peak
from visually crushing the real case counts in earlier years.

```{r ts-overview, fig.height=11, fig.cap="Figure 1: Monthly time series of dengue cases and meteorological variables (2008–2025). Red dashed line = LOESS trend. Dengue panel on log1p scale to reveal early-year case counts."}

# ── Panel A: Dengue — log1p scale ─────────────────────────────────────────────
p_dengue_ts <- df_raw %>%
  dplyr::select(Date, DengueCases) %>%
  ggplot(aes(x = Date, y = DengueCases + 1)) +
  geom_line(colour = "#2C7BB6", linewidth = 0.65) +
  geom_smooth(method = "loess", span = 0.25, se = FALSE,
              colour = "#D7191C", linewidth = 0.9, linetype = "dashed") +
  scale_x_date(date_labels = "%Y", date_breaks = "2 years",
               limits = c(as.Date("2008-01-01"), NA)) +
  scale_y_continuous(
    trans  = "log1p",
    breaks = c(1, 11, 101, 1001, 10001, 100001),
    labels = c("0", "10", "100", "1,000", "10,000", "100,000")
  ) +
  labs(x = NULL, y = "Cases") +
  ggtitle("Dengue Cases (count)  \u2014  log\u2081\u209a scale") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 10),
    strip.background   = element_rect(fill = "#EEF3F8", colour = NA),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line(colour = "grey88"),
    axis.text.x        = element_text(angle = 45, hjust = 1, size = 9)
  )

# ── Panel B: Climate variables — linear scale ─────────────────────────────────
p_climate_ts <- df_raw %>%
  dplyr::select(Date, MeanTemp, Precipitation, Humidity, WindSpeed) %>%
  tidyr::pivot_longer(-Date, names_to = "Variable", values_to = "Value") %>%
  dplyr::mutate(Variable = factor(Variable,
    levels = c("MeanTemp", "Precipitation", "Humidity", "WindSpeed"),
    labels = c("Mean Temperature (\u00b0C)", "Precipitation (mm)",
               "Relative Humidity (%)", "Wind Speed (m/s)"))) %>%
  ggplot(aes(x = Date, y = Value)) +
  geom_line(colour = "#2C7BB6", linewidth = 0.65) +
  geom_smooth(method = "loess", span = 0.25, se = FALSE,
              colour = "#D7191C", linewidth = 0.9, linetype = "dashed") +
  facet_wrap(~ Variable, scales = "free_y", ncol = 1) +
  scale_x_date(date_labels = "%Y", date_breaks = "2 years",
               limits = c(as.Date("2008-01-01"), NA)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text         = element_text(face = "bold", size = 10),
    strip.background   = element_rect(fill = "#EEF3F8", colour = NA),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line(colour = "grey88"),
    axis.text.x        = element_text(angle = 45, hjust = 1, size = 9)
  )

# ── Combine: dengue ~20% height, climate ~80% ─────────────────────────────────
gridExtra::grid.arrange(
  p_dengue_ts,
  p_climate_ts,
  nrow    = 2,
  heights = grid::unit(c(1.2, 4), "null")
)
```


### 2.3.2 Climate–Dengue Bivariate Scatter Plots

Pairwise associations between each meteorological predictor and dengue counts
($\log_{10}$ scale; zero-case months excluded for legibility). LOESS curves reveal the
non-linear exposure-response shapes that motivate the distributed-lag modelling
framework in Stage 1.

```{r climate-dengue-scatter, fig.height=7, fig.cap="Figure 2: Climate variable vs. dengue cases (non-zero months). LOESS smoother with 95% CI; y-axis log10 scale."}
df_scatter_long <- df_raw %>%
  dplyr::filter(DengueCases > 0) %>%
  dplyr::select(DengueCases, MeanTemp, Precipitation, Humidity, WindSpeed) %>%
  tidyr::pivot_longer(-DengueCases, names_to = "Climate_Var", values_to = "Climate_Val") %>%
  dplyr::mutate(Climate_Var = factor(Climate_Var,
    levels = c("MeanTemp", "Precipitation", "Humidity", "WindSpeed"),
    labels = c("Mean Temperature (\u00b0C)", "Precipitation (mm)",
               "Relative Humidity (%)", "Wind Speed (m/s)")))

ggplot(df_scatter_long, aes(x = Climate_Val, y = DengueCases)) +
  geom_point(colour = "#2C7BB6", alpha = 0.55, size = 2.0) +
  geom_smooth(method    = "loess", span = 0.7,
              se        = TRUE,
              colour    = "#D7191C",
              fill      = "#D7191C",
              alpha     = 0.15,
              linewidth = 1.0) +
  scale_y_log10(labels = scales::comma,
                breaks  = c(1, 10, 100, 1000, 10000, 100000)) +
  facet_wrap(~ Climate_Var, scales = "free_x", ncol = 2) +
  labs(
    x = NULL,
    y = expression("Dengue Cases (" * log[10] * " scale)")
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "#EEF3F8", colour = NA),
    panel.grid.minor = element_blank()
  )
```

---

# 3. Global Settings

```{r global-settings}
lag_period <- 6    # maximum lag window (months); matches A. aegypti full cycle
max_lags   <- 12   # search range for autocorrelation correction lags
trend_df   <- 5    # ns() df for long-term trend spline

# Prediction range: 5th–95th percentile (avoids spline boundary artefacts)
pred_MeanTemp_range  <- quantile(df_raw$MeanTemp,      c(0.05, 0.95))
pred_Precip_range    <- quantile(df_raw$Precipitation, c(0.05, 0.95))
pred_Humidity_range  <- quantile(df_raw$Humidity,      c(0.05, 0.95))
pred_WindSpeed_range <- quantile(df_raw$WindSpeed,     c(0.05, 0.95))

# Reference values: median is used (more robust for right-skewed data)
ref_MeanTemp  <- median(df_raw$MeanTemp)
ref_Precip    <- median(df_raw$Precipitation)
ref_Humidity  <- median(df_raw$Humidity)
ref_WindSpeed <- median(df_raw$WindSpeed)

cat("Prediction ranges (5th–95th percentile):\n")
cat("  MeanTemp:     ", round(pred_MeanTemp_range,  2), "°C\n")
cat("  Precipitation:", round(pred_Precip_range,    2), "mm\n")
cat("  Humidity:     ", round(pred_Humidity_range,  2), "%\n")
cat("  WindSpeed:    ", round(pred_WindSpeed_range, 2), "m/s\n")
cat("\nReference values (medians):\n")
cat("  MeanTemp:", round(ref_MeanTemp, 2), "°C |",
    "Precip:", round(ref_Precip, 2), "mm |",
    "Humidity:", round(ref_Humidity, 2), "% |",
    "WindSpeed:", round(ref_WindSpeed, 2), "m/s\n")

# Helper: preserve dlnm class/attributes when row-subsetting a crossbasis matrix
subset_cb <- function(cb, rows) {
  cb_sub <- cb[rows, , drop = FALSE]
  for (a in setdiff(names(attributes(cb)), c("dim", "dimnames")))
    attr(cb_sub, a) <- attr(cb, a)
  class(cb_sub) <- class(cb)
  cb_sub
}
```

---

# 4. Stage 1 — DLNM Epidemiological Characterisation

Cross-bases are built on the **full 216-row dataset** to ensure every observation
has a complete 6-month lag history. The first `lag_period` rows are then trimmed
before model fitting (rows with incomplete lag windows).

## 4.1 Cross-Basis Construction

```{r build-crossbases}
cb_MT_full <- crossbasis(df_raw$MeanTemp, lag = lag_period,
  argvar = list(fun = "ns", df = 2, Boundary.knots = pred_MeanTemp_range),
  arglag = list(fun = "ns", df = 2))

cb_PR_full <- crossbasis(df_raw$Precipitation, lag = lag_period,
  argvar = list(fun = "ns", df = 3, Boundary.knots = pred_Precip_range),
  arglag = list(fun = "ns", df = 3))

cb_HU_full <- crossbasis(df_raw$Humidity, lag = lag_period,
  argvar = list(fun = "ns", df = 3, Boundary.knots = pred_Humidity_range),
  arglag = list(fun = "ns", df = 3))

# Wind speed: full observed range as boundary knots (22 obs outside 5th–95th pct)
cb_WS_full <- crossbasis(df_raw$WindSpeed, lag = lag_period,
  argvar = list(fun = "ns", df = 2, Boundary.knots = range(df_raw$WindSpeed)),
  arglag = list(fun = "ns", df = 2))

fit_rows <- (lag_period + 1):nrow(df_raw)
df       <- df_raw[fit_rows, ]
cb_MT    <- subset_cb(cb_MT_full, fit_rows)
cb_PR    <- subset_cb(cb_PR_full, fit_rows)
cb_HU    <- subset_cb(cb_HU_full, fit_rows)
cb_WS    <- subset_cb(cb_WS_full, fit_rows)

cat("Fitting rows:", nrow(df),
    "| NA check — MT:", sum(is.na(cb_MT)),
    "PR:", sum(is.na(cb_PR)),
    "HU:", sum(is.na(cb_HU)),
    "WS:", sum(is.na(cb_WS)), "\n")
```

## 4.2 Autocorrelation Correction (Durbin-Watson grid search)

For each variable: fit a penultimate model, extract deviance residuals, then search
over k = 1–12 lagged residual terms selecting the k that minimises |DW_k − 2|.

```{r dw-helper}
# Generic DW grid search: returns best_k and vectors of DW / AIC per k
dw_grid_search <- function(cb_full, confounders_str, df_all, fit_rows,
                            res_pent, max_lags, label) {
  dw_vec  <- numeric(max_lags)
  aic_vec <- numeric(max_lags)

  for (n_lags in seq_len(max_lags)) {
    df_t <- df_all
    for (i in seq_len(n_lags))
      df_t[[paste0("lr", i)]] <- dplyr::lag(res_pent, i)

    cb_t     <- subset_cb(cb_full, fit_rows)
    lag_str  <- paste0("lr", seq_len(n_lags), collapse = " + ")
    fml <- as.formula(paste(
      "DengueCases ~ ns(Month_index, df=trend_df) + cb_t +",
      confounders_str, "+", lag_str,
      "+ factor(Month_num) + offset(log(Population))"
    ))
    m <- tryCatch(glm.nb(fml, data = df_t, na.action = na.omit),
                  error = function(e) NULL)
    if (is.null(m)) { dw_vec[n_lags] <- NA; aic_vec[n_lags] <- NA; next }
    aic_vec[n_lags] <- AIC(m)
    dw_vec[n_lags]  <- dwtest(lm(residuals(m, type = "deviance") ~ 1))$statistic
  }

  vi       <- which(!is.na(dw_vec))
  best_k   <- vi[which.min(abs(dw_vec[vi] - 2))]
  cat(sprintf("  %-15s best_k = %d | DW = %.3f | AIC = %.1f\n",
              label, best_k, dw_vec[best_k], aic_vec[best_k]))
  list(best_k = best_k, dw = dw_vec, aic = aic_vec)
}
```

## 4.3 Model 1 — Mean Temperature

```{r dlnm-temperature}
# Penultimate model (no lagged residuals yet)
MT_penultimate <- glm.nb(
  DengueCases ~ ns(Month_index, df = trend_df) + cb_MT +
    Precipitation + Humidity + WindSpeed +
    factor(Month_num) + offset(log(Population)),
  data = df
)
res_MT <- residuals(MT_penultimate, type = "deviance")

dw_MT   <- dw_grid_search(cb_MT_full, "Precipitation + Humidity + WindSpeed",
                           df, fit_rows, res_MT, max_lags, "MeanTemp")
best_MT <- dw_MT$best_k

# Final model
df_MT <- df
for (i in seq_len(best_MT)) df_MT[[paste0("lr", i)]] <- dplyr::lag(res_MT, i)
complete_MT <- complete.cases(df_MT)
df_MT       <- df_MT[complete_MT, ]
cb_MT       <- subset_cb(cb_MT_full, fit_rows[complete_MT])

fml_MT <- as.formula(paste(
  "DengueCases ~ ns(Month_index, df=trend_df) + cb_MT +",
  "Precipitation + Humidity + WindSpeed +",
  paste0("lr", seq_len(best_MT), collapse = " + "),
  "+ factor(Month_num) + offset(log(Population))"
))
model_MT <- glm.nb(fml_MT, data = df_MT)

null_MT  <- glm.nb(DengueCases ~ 1 + offset(log(Population)), data = df_MT)
ll_diff  <- 2 * (as.numeric(logLik(model_MT)) - as.numeric(logLik(null_MT)))
df_diff  <- attr(logLik(model_MT), "df") - attr(logLik(null_MT), "df")
cat(sprintf("MeanTemp — AIC: %.1f | DW: %.3f | LR chi2: %.1f (df=%d, p%s)\n",
    AIC(model_MT),
    dwtest(lm(residuals(model_MT, type="deviance") ~ 1))$statistic,
    ll_diff, df_diff,
    ifelse(pchisq(ll_diff, df_diff, lower.tail=FALSE) < 0.001, " < 0.001",
           paste0(" = ", round(pchisq(ll_diff, df_diff, lower.tail=FALSE), 4)))))

# Exposure-lag-response surface
pred_MT <- crosspred(cb_MT, model_MT,
  at       = seq(pred_MeanTemp_range[1], pred_MeanTemp_range[2], by = 0.1),
  cumul    = TRUE, cen = ref_MeanTemp, ci.level = 0.95, vcov = vcov(model_MT))

plot(pred_MT, "3d", xlab = "Mean Temp (°C)", ylab = "Lag (months)", zlab = "RR",
     col = "white", shade = 0.6, theta = -145, phi = 38,
     main = "Mean Temperature: Exposure-Lag-Response Surface")
plot(pred_MT, "overall", xlab = "Mean Temperature (°C)", ylab = "Cumulative RR",
     main = "Mean Temperature: Cumulative Exposure-Response",
     lwd = 2, col = "darkred",
     ci.arg = list(col = adjustcolor("red", alpha.f = 0.2)))
abline(h = 1, lty = 2, col = "grey40"); grid()

# Lag-specific slices
par(mfrow = c(2, ceiling((lag_period + 1) / 2)))
for (i in 0:lag_period)
  plot(pred_MT, "slices", lag = i, ci = "area",
       xlab = "Mean Temp (°C)", ylab = "RR", main = paste("Lag", i, "month(s)"))
par(mfrow = c(1, 1))
```

## 4.4 Model 2 — Precipitation

```{r dlnm-precipitation}
PR_penultimate <- glm.nb(
  DengueCases ~ ns(Month_index, df = trend_df) + cb_PR +
    MeanTemp + Humidity + WindSpeed +
    factor(Month_num) + offset(log(Population)),
  data = df
)
res_PR <- residuals(PR_penultimate, type = "deviance")

dw_PR   <- dw_grid_search(cb_PR_full, "MeanTemp + Humidity + WindSpeed",
                           df, fit_rows, res_PR, max_lags, "Precipitation")
best_PR <- dw_PR$best_k

df_PR <- df
for (i in seq_len(best_PR)) df_PR[[paste0("lr", i)]] <- dplyr::lag(res_PR, i)
complete_PR <- complete.cases(df_PR)
df_PR       <- df_PR[complete_PR, ]
cb_PR       <- subset_cb(cb_PR_full, fit_rows[complete_PR])

fml_PR <- as.formula(paste(
  "DengueCases ~ ns(Month_index, df=trend_df) + cb_PR +",
  "MeanTemp + Humidity + WindSpeed +",
  paste0("lr", seq_len(best_PR), collapse = " + "),
  "+ factor(Month_num) + offset(log(Population))"
))
model_PR <- glm.nb(fml_PR, data = df_PR)

null_PR  <- glm.nb(DengueCases ~ 1 + offset(log(Population)), data = df_PR)
ll_diff  <- 2 * (as.numeric(logLik(model_PR)) - as.numeric(logLik(null_PR)))
df_diff  <- attr(logLik(model_PR), "df") - attr(logLik(null_PR), "df")
cat(sprintf("Precipitation — AIC: %.1f | DW: %.3f | LR chi2: %.1f (df=%d, p%s)\n",
    AIC(model_PR),
    dwtest(lm(residuals(model_PR, type="deviance") ~ 1))$statistic,
    ll_diff, df_diff,
    ifelse(pchisq(ll_diff, df_diff, lower.tail=FALSE) < 0.001, " < 0.001",
           paste0(" = ", round(pchisq(ll_diff, df_diff, lower.tail=FALSE), 4)))))

pred_PR <- crosspred(cb_PR, model_PR,
  at       = seq(pred_Precip_range[1], pred_Precip_range[2], by = 1),
  cumul    = TRUE, cen = ref_Precip, ci.level = 0.95, vcov = vcov(model_PR))

plot(pred_PR, "3d", xlab = "Precipitation (mm)", ylab = "Lag (months)", zlab = "RR",
     col = "white", shade = 0.6, theta = -145, phi = 38,
     main = "Precipitation: Exposure-Lag-Response Surface")
plot(pred_PR, "overall", xlab = "Precipitation (mm)", ylab = "Cumulative RR",
     main = "Precipitation: Cumulative Exposure-Response",
     lwd = 2, col = "darkblue",
     ci.arg = list(col = adjustcolor("blue", alpha.f = 0.2)))
abline(h = 1, lty = 2, col = "grey40"); grid()

par(mfrow = c(2, ceiling((lag_period + 1) / 2)))
for (i in 0:lag_period)
  plot(pred_PR, "slices", lag = i, ci = "area",
       xlab = "Precipitation (mm)", ylab = "RR", main = paste("Lag", i, "month(s)"))
par(mfrow = c(1, 1))
```

## 4.5 Model 3 — Relative Humidity

```{r dlnm-humidity}
HU_penultimate <- glm.nb(
  DengueCases ~ ns(Month_index, df = trend_df) + cb_HU +
    MeanTemp + Precipitation + WindSpeed +
    factor(Month_num) + offset(log(Population)),
  data = df
)
res_HU <- residuals(HU_penultimate, type = "deviance")

dw_HU   <- dw_grid_search(cb_HU_full, "MeanTemp + Precipitation + WindSpeed",
                           df, fit_rows, res_HU, max_lags, "Humidity")
best_HU <- dw_HU$best_k

df_HU <- df
for (i in seq_len(best_HU)) df_HU[[paste0("lr", i)]] <- dplyr::lag(res_HU, i)
complete_HU <- complete.cases(df_HU)
df_HU       <- df_HU[complete_HU, ]
cb_HU       <- subset_cb(cb_HU_full, fit_rows[complete_HU])

fml_HU <- as.formula(paste(
  "DengueCases ~ ns(Month_index, df=trend_df) + cb_HU +",
  "MeanTemp + Precipitation + WindSpeed +",
  paste0("lr", seq_len(best_HU), collapse = " + "),
  "+ factor(Month_num) + offset(log(Population))"
))
model_HU <- glm.nb(fml_HU, data = df_HU)

null_HU  <- glm.nb(DengueCases ~ 1 + offset(log(Population)), data = df_HU)
ll_diff  <- 2 * (as.numeric(logLik(model_HU)) - as.numeric(logLik(null_HU)))
df_diff  <- attr(logLik(model_HU), "df") - attr(logLik(null_HU), "df")
cat(sprintf("Humidity — AIC: %.1f | DW: %.3f | LR chi2: %.1f (df=%d, p%s)\n",
    AIC(model_HU),
    dwtest(lm(residuals(model_HU, type="deviance") ~ 1))$statistic,
    ll_diff, df_diff,
    ifelse(pchisq(ll_diff, df_diff, lower.tail=FALSE) < 0.001, " < 0.001",
           paste0(" = ", round(pchisq(ll_diff, df_diff, lower.tail=FALSE), 4)))))

pred_HU <- crosspred(cb_HU, model_HU,
  at       = seq(pred_Humidity_range[1], pred_Humidity_range[2], by = 0.5),
  cumul    = TRUE, cen = ref_Humidity, ci.level = 0.95, vcov = vcov(model_HU))

plot(pred_HU, "3d", xlab = "Humidity (%)", ylab = "Lag (months)", zlab = "RR",
     col = "white", shade = 0.6, theta = -145, phi = 38,
     main = "Humidity: Exposure-Lag-Response Surface")
plot(pred_HU, "overall", xlab = "Humidity (%)", ylab = "Cumulative RR",
     main = "Humidity: Cumulative Exposure-Response",
     lwd = 2, col = "darkgreen",
     ci.arg = list(col = adjustcolor("green", alpha.f = 0.2)))
abline(h = 1, lty = 2, col = "grey40"); grid()

par(mfrow = c(2, ceiling((lag_period + 1) / 2)))
for (i in 0:lag_period)
  plot(pred_HU, "slices", lag = i, ci = "area",
       xlab = "Humidity (%)", ylab = "RR", main = paste("Lag", i, "month(s)"))
par(mfrow = c(1, 1))
```

## 4.6 Model 4 — Wind Speed

Wind speed requires Poisson warm-start due to NB convergence issues at this
temporal resolution.

```{r dlnm-windspeed}
# Rebuild WS crossbasis with full-range boundary knots
cb_WS_full <- crossbasis(df_raw$WindSpeed, lag = lag_period,
  argvar = list(fun = "ns", df = 2, Boundary.knots = range(df_raw$WindSpeed)),
  arglag = list(fun = "ns", df = 2))
cb_WS <- subset_cb(cb_WS_full, fit_rows)

# Poisson warm-start for NB convergence
WS_pois <- glm(
  DengueCases ~ ns(Month_index, df = trend_df) + cb_WS +
    MeanTemp + Precipitation + Humidity +
    factor(Month_num) + offset(log(Population)),
  data = df, family = poisson, control = glm.control(maxit = 200)
)
WS_penultimate <- glm.nb(
  DengueCases ~ ns(Month_index, df = trend_df) + cb_WS +
    MeanTemp + Precipitation + Humidity +
    factor(Month_num) + offset(log(Population)),
  data = df, start = coef(WS_pois),
  control = glm.control(maxit = 500, epsilon = 1e-6)
)
res_WS <- residuals(WS_penultimate, type = "deviance")

# DW grid search with Poisson warm-start per fold
dw_WS_vec  <- numeric(max_lags)
aic_WS_vec <- numeric(max_lags)
for (n_lags in seq_len(max_lags)) {
  df_t <- df
  for (i in seq_len(n_lags)) df_t[[paste0("lr", i)]] <- dplyr::lag(res_WS, i)
  cb_t     <- subset_cb(cb_WS_full, fit_rows)
  lag_str  <- paste0("lr", seq_len(n_lags), collapse = " + ")
  fml <- as.formula(paste(
    "DengueCases ~ ns(Month_index, df=trend_df) + cb_t +",
    "MeanTemp + Precipitation + Humidity +", lag_str,
    "+ factor(Month_num) + offset(log(Population))"
  ))
  m <- tryCatch({
    p <- glm(fml, data = df_t, family = poisson,
             control = glm.control(maxit = 200), na.action = na.omit)
    glm.nb(fml, data = df_t, start = coef(p),
           control = glm.control(maxit = 500, epsilon = 1e-6), na.action = na.omit)
  }, error = function(e) NULL)
  if (is.null(m)) { dw_WS_vec[n_lags] <- NA; aic_WS_vec[n_lags] <- NA; next }
  aic_WS_vec[n_lags] <- AIC(m)
  dw_WS_vec[n_lags]  <- dwtest(lm(residuals(m, type = "deviance") ~ 1))$statistic
}

vi       <- which(!is.na(dw_WS_vec))
best_WS  <- vi[which.min(abs(dw_WS_vec[vi] - 2))]
cat(sprintf("  %-15s best_k = %d | DW = %.3f\n", "WindSpeed",
            best_WS, dw_WS_vec[best_WS]))

df_WS <- df
for (i in seq_len(best_WS)) df_WS[[paste0("lr", i)]] <- dplyr::lag(res_WS, i)
complete_WS <- complete.cases(df_WS)
df_WS       <- df_WS[complete_WS, ]
cb_WS       <- subset_cb(cb_WS_full, fit_rows[complete_WS])

fml_WS <- as.formula(paste(
  "DengueCases ~ ns(Month_index, df=trend_df) + cb_WS +",
  "MeanTemp + Precipitation + Humidity +",
  paste0("lr", seq_len(best_WS), collapse = " + "),
  "+ factor(Month_num) + offset(log(Population))"
))
WS_pois_f <- glm(fml_WS, data = df_WS, family = poisson,
                 control = glm.control(maxit = 200))
model_WS  <- glm.nb(fml_WS, data = df_WS,
                    start = coef(WS_pois_f),
                    control = glm.control(maxit = 500, epsilon = 1e-6))

null_WS  <- glm.nb(DengueCases ~ 1 + offset(log(Population)), data = df_WS)
ll_diff  <- 2 * (as.numeric(logLik(model_WS)) - as.numeric(logLik(null_WS)))
df_diff  <- attr(logLik(model_WS), "df") - attr(logLik(null_WS), "df")
cat(sprintf("WindSpeed — AIC: %.1f | DW: %.3f | LR chi2: %.1f (df=%d, p%s)\n",
    AIC(model_WS),
    dwtest(lm(residuals(model_WS, type="deviance") ~ 1))$statistic,
    ll_diff, df_diff,
    ifelse(pchisq(ll_diff, df_diff, lower.tail=FALSE) < 0.001, " < 0.001",
           paste0(" = ", round(pchisq(ll_diff, df_diff, lower.tail=FALSE), 4)))))

pred_WS <- crosspred(cb_WS, model_WS,
  at       = seq(pred_WindSpeed_range[1], pred_WindSpeed_range[2], by = 0.05),
  cumul    = TRUE, cen = ref_WindSpeed, ci.level = 0.95, vcov = vcov(model_WS))

plot(pred_WS, "3d", xlab = "Wind Speed (m/s)", ylab = "Lag (months)", zlab = "RR",
     col = "white", shade = 0.6, theta = -145, phi = 38,
     main = "Wind Speed: Exposure-Lag-Response Surface")
plot(pred_WS, "overall", xlab = "Wind Speed (m/s)", ylab = "Cumulative RR",
     main = "Wind Speed: Cumulative Exposure-Response",
     lwd = 2, col = "purple",
     ci.arg = list(col = adjustcolor("purple", alpha.f = 0.2)))
abline(h = 1, lty = 2, col = "grey40"); grid()

par(mfrow = c(2, ceiling((lag_period + 1) / 2)))
for (i in 0:lag_period)
  plot(pred_WS, "slices", lag = i, ci = "area",
    xlab = "Wind Speed (m/s)", ylab = "RR", main = paste("Lag", i, "month(s)"))
par(mfrow = c(1, 1))
```

## 4.7 DLNM Summary

```{r dlnm-summary}
cat("=== DLNM Model Summary ===\n")
model_map <- list(
  MeanTemp = list(model = model_MT, best = best_MT),
  Precipitation = list(model = model_PR, best = best_PR),
  Humidity = list(model = model_HU, best = best_HU),
  WindSpeed = list(model = model_WS, best = best_WS)
)
for (nm in names(model_map)) {
  mod  <- model_map[[nm]]$model
  best <- model_map[[nm]]$best
  dw_f <- dwtest(lm(residuals(mod, type="deviance") ~ 1))$statistic
  cat(sprintf("  %-15s | AIC: %8.1f | DW: %.3f | Residual lags: %d\n",
              nm, AIC(mod), dw_f, best))
}
```

---

# 5. Stage 2 — CBPFE: Leakage-Free Feature Extraction

DLNM models are **refit on the training partition only** (Year ≤ 2023).
Cross-basis partial effects for all rows are then computed via deterministic
matrix multiplication: `f_t = B_t × β̂`, where `β̂` comes from training data only.

```{r cbpfe-feature-extraction}
# Align CB matrices with df (210 rows)
cb_MT_ml <- subset_cb(cb_MT_full, fit_rows)
cb_PR_ml <- subset_cb(cb_PR_full, fit_rows)
cb_HU_ml <- subset_cb(cb_HU_full, fit_rows)
cb_WS_ml <- subset_cb(cb_WS_full, fit_rows)

train_idx_dlnm <- which(df$Year <= 2023)
cat("Refitting DLNM on training rows:", length(train_idx_dlnm), "(2008–2023)\n")

# Fit training-only DLNM and project partial effects to all rows
fit_train_dlnm <- function(cb_ml, df_all, train_idx, confounders, label) {
  cb_tr <- subset_cb(cb_ml, train_idx)
  df_tr <- df_all[train_idx, ]
  .GlobalEnv$cb_oos <- cb_tr

  fml <- as.formula(paste0(
    "DengueCases ~ ns(Month_index, df=", trend_df, ") + cb_oos + ",
    confounders, " + factor(Month_num) + offset(log(Population))"
  ))
  pois <- tryCatch(
    glm(fml, data = df_tr, family = poisson, control = glm.control(maxit = 200)),
    error = function(e) NULL)
  nb <- tryCatch({
    if (!is.null(pois))
      glm.nb(fml, data = df_tr, start = coef(pois),
             control = glm.control(maxit = 500, epsilon = 1e-6))
    else
      glm.nb(fml, data = df_tr, control = glm.control(maxit = 500))
  }, error = function(e) {
    warning(paste(label, "NB failed; using Poisson")); pois
  })

  # Step 3 of Algorithm 1: deterministic projection (no future y_t used)
  cf       <- coef(nb)
  cb_coefs <- cf[grep("^cb_oos", names(cf))]
  effect   <- as.numeric(cb_ml %*% cb_coefs)
  cat(sprintf("  %-15s coefs=%d | effect [%+.3f, %+.3f]\n",
              label, length(cb_coefs), min(effect), max(effect)))
  list(model = nb, effect = effect)
}

dlnm_MT <- fit_train_dlnm(cb_MT_ml, df, train_idx_dlnm,
                           "Precipitation + Humidity + WindSpeed", "MeanTemp")
dlnm_PR <- fit_train_dlnm(cb_PR_ml, df, train_idx_dlnm,
                           "MeanTemp + Humidity + WindSpeed",       "Precipitation")
dlnm_HU <- fit_train_dlnm(cb_HU_ml, df, train_idx_dlnm,
                           "MeanTemp + Precipitation + WindSpeed",  "Humidity")
dlnm_WS <- fit_train_dlnm(cb_WS_ml, df, train_idx_dlnm,
                           "MeanTemp + Precipitation + Humidity",   "WindSpeed")
rm(cb_oos, envir = .GlobalEnv)

df$dlnm_effect_MT <- dlnm_MT$effect
df$dlnm_effect_PR <- dlnm_PR$effect
df$dlnm_effect_HU <- dlnm_HU$effect
df$dlnm_effect_WS <- dlnm_WS$effect
dlnm_effect_cols  <- c("dlnm_effect_MT", "dlnm_effect_PR",
                        "dlnm_effect_HU", "dlnm_effect_WS")
cat("\nAll DLNM features are leak-free (trained on 2008–2023 only)\n")
```

---

# 6. Feature Assembly and Train/Test Split

```{r feature-assembly}
df_ml <- df

# AR lags on log1p scale (consistent with log-incidence target)
for (k in c(1:6, 12))
  df_ml[[paste0("lag", k, "_dengue")]] <- dplyr::lag(log1p(df_ml$DengueCases), k)
ar_cols     <- paste0("lag", c(1:6, 12), "_dengue")

# Cyclical month encoding
df_ml$sin_month <- sin(2 * pi * df_ml$Month_num / 12)
df_ml$cos_month <- cos(2 * pi * df_ml$Month_num / 12)
season_cols     <- c("sin_month", "cos_month")

# Target: log-incidence per 100,000 population
df_ml$log_incidence <- log1p(df_ml$DengueCases / df_ml$Population * 1e5)

needed_cols <- c(ar_cols, "log_incidence", dlnm_effect_cols)
train_df <- df_ml %>% filter(Year <= 2023) %>% tidyr::drop_na(all_of(needed_cols))
test_df  <- df_ml %>% filter(Year >= 2024) %>% tidyr::drop_na(all_of(needed_cols))

cat("Train:", nrow(train_df), "rows |",
    format(min(train_df$Date)), "–", format(max(train_df$Date)), "\n")
cat("Test: ", nrow(test_df),  "rows |",
    format(min(test_df$Date)),  "–", format(max(test_df$Date)), "\n")

# Three feature sets (Step 4 of Algorithm 1)
feat_a <- c(ar_cols, season_cols)                          # Model A: AR only
feat_b <- c(ar_cols, season_cols, dlnm_effect_cols)        # Model B: AR + CBPFE
ar_reduced <- paste0("lag", c(1, 2, 6, 12), "_dengue")
feat_c <- c(ar_reduced, season_cols, dlnm_effect_cols)     # Model C: hybrid

cat("\nFeature counts — A:", length(feat_a),
    "| B:", length(feat_b), "| C:", length(feat_c), "\n")
```

---

# 7. Helper Functions

```{r helpers}
# Back-transform log-incidence to absolute case counts
bt <- function(log_inc, pop) pmax(0, expm1(log_inc) / 1e5 * pop)

# RMSE / MAE / R2
calc_metrics <- function(actual, pred) {
  list(rmse = sqrt(mean((actual - pred)^2)),
       mae  = mean(abs(actual - pred)),
       r2   = 1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2))
}

# Time-series CV fold generator
generate_cv_folds <- function(n, min_train, horizon, step, type = "expanding") {
  folds <- list()
  if (type == "expanding") {
    for (end in seq(min_train, n - horizon, by = step))
      folds[[length(folds) + 1]] <- list(
        train = seq_len(end),
        val   = seq.int(end + 1, min(end + horizon, n)))
  } else {
    for (start in seq(1, n - min_train - horizon + 1, by = step)) {
      tr_end <- start + min_train - 1
      va_end <- min(tr_end + horizon, n)
      if (va_end > n) break
      folds[[length(folds) + 1]] <- list(
        train = seq.int(start, tr_end),
        val   = seq.int(tr_end + 1, va_end))
    }
  }
  folds
}

# Run CV for a given parameter set
run_cv <- function(train_data, feat_cols, params, folds) {
  x_full <- as.matrix(train_data[, feat_cols])
  y_full <- train_data$log_incidence
  rmse_v <- numeric(length(folds))
  nr_v   <- integer(length(folds))
  for (i in seq_along(folds)) {
    tr <- folds[[i]]$train;  va <- folds[[i]]$val
    dtr <- xgb.DMatrix(data = x_full[tr, , drop = FALSE], label = y_full[tr])
    dva <- xgb.DMatrix(data = x_full[va, , drop = FALSE], label = y_full[va])
    set.seed(42)
    suppressWarnings(
      m <- xgb.train(params = params, data = dtr, nrounds = 300,
                     watchlist = list(val = dva),
                     early_stopping_rounds = 30, verbose = 0))
    pred_c      <- bt(predict(m, dva), train_data$Population[va])
    rmse_v[i]   <- sqrt(mean((train_data$DengueCases[va] - pred_c)^2))
    nr_v[i]     <- if (!is.null(m$best_iteration) && m$best_iteration > 0)
                     m$best_iteration else 300L
  }
  list(mean_rmse = mean(rmse_v), sd_rmse = sd(rmse_v),
       rmse_vec = rmse_v, nrounds_vec = nr_v,
       median_nrounds = as.integer(median(nr_v)),
       n_folds = length(folds))
}
```

---

# 8. Expanding vs Rolling CV Comparison

```{r cv-comparison}
n_train   <- nrow(train_df)
min_train <- 48;  horizon_cv <- 12;  step_cv <- 6

folds_expanding <- generate_cv_folds(n_train, min_train, horizon_cv, step_cv, "expanding")
folds_rolling   <- generate_cv_folds(n_train, min_train, horizon_cv, step_cv, "rolling")

default_params <- list(
  objective = "reg:squarederror", max_depth = 4, eta = 0.05,
  subsample = 0.8, colsample_bytree = 0.8,
  min_child_weight = 5, gamma = 1, lambda = 5
)

cv_exp <- run_cv(train_df, feat_c, default_params, folds_expanding)
cv_rol <- run_cv(train_df, feat_c, default_params, folds_rolling)

cat(sprintf("Expanding  | folds=%d | mean RMSE=%.0f | SD=%.0f\n",
            cv_exp$n_folds, cv_exp$mean_rmse, cv_exp$sd_rmse))
cat(sprintf("Rolling    | folds=%d | mean RMSE=%.0f | SD=%.0f\n",
            cv_rol$n_folds, cv_rol$mean_rmse, cv_rol$sd_rmse))

if (cv_exp$mean_rmse <= cv_rol$mean_rmse) {
  best_cv_type  <- "expanding"; best_cv_folds <- folds_expanding
} else {
  best_cv_type  <- "rolling";   best_cv_folds <- folds_rolling
}
cat(">>> Selected:", toupper(best_cv_type), "CV\n")
```

---

# 9. Hyperparameter Tuning (Random Grid Search, n=20)

```{r hyperparameter-tuning}
set.seed(123)
n_grid <- 20
param_grid <- data.frame(
  max_depth        = sample(c(3,4,5,6),        n_grid, replace = TRUE),
  eta              = sample(c(0.01,0.03,0.05,0.1), n_grid, replace = TRUE),
  subsample        = sample(c(0.7,0.8,0.9),    n_grid, replace = TRUE),
  colsample_bytree = sample(c(0.6,0.8,1.0),    n_grid, replace = TRUE),
  min_child_weight = sample(c(3,5,10,15),       n_grid, replace = TRUE),
  gamma            = sample(c(0,0.5,1,3),       n_grid, replace = TRUE),
  lambda           = sample(c(1,3,5,10),        n_grid, replace = TRUE)
)

tune_model <- function(train_data, feat_cols, folds, label) {
  cat(sprintf("Tuning %s (%d features)...\n", label, length(feat_cols)))
  results <- data.frame()
  for (j in seq_len(n_grid)) {
    p_j <- list(objective="reg:squarederror",
                max_depth=param_grid$max_depth[j], eta=param_grid$eta[j],
                subsample=param_grid$subsample[j],
                colsample_bytree=param_grid$colsample_bytree[j],
                min_child_weight=param_grid$min_child_weight[j],
                gamma=param_grid$gamma[j], lambda=param_grid$lambda[j])
    cv_r <- run_cv(train_data, feat_cols, p_j, folds)
    results <- rbind(results, cbind(param_grid[j, ],
                     mean_rmse = cv_r$mean_rmse, sd_rmse = cv_r$sd_rmse,
                     median_nrounds = cv_r$median_nrounds))
  }
  best_j <- which.min(results$mean_rmse)
  cat(sprintf("  Best: RMSE=%.0f | depth=%d eta=%.3f gamma=%.1f lambda=%d nrounds=%d\n",
              results$mean_rmse[best_j], results$max_depth[best_j],
              results$eta[best_j], results$gamma[best_j],
              results$lambda[best_j], results$median_nrounds[best_j]))
  list(
    best_params = list(objective="reg:squarederror",
      max_depth=results$max_depth[best_j], eta=results$eta[best_j],
      subsample=results$subsample[best_j],
      colsample_bytree=results$colsample_bytree[best_j],
      min_child_weight=results$min_child_weight[best_j],
      gamma=results$gamma[best_j], lambda=results$lambda[best_j]),
    best_nrounds = results$median_nrounds[best_j],
    all_results  = results[order(results$mean_rmse), ]
  )
}

tuned_a <- tune_model(train_df, feat_a, best_cv_folds, "Model A: AR only")
tuned_b <- tune_model(train_df, feat_b, best_cv_folds, "Model B: AR + CBPFE")
tuned_c <- tune_model(train_df, feat_c, best_cv_folds, "Model C: reduced AR + CBPFE")

# Top-3 stability check (confirms hyperparameter sensitivity is low)
cat("\n=== Top-3 Configurations per Model ===\n")
for (nm in c("a","b","c")) {
  cat(sprintf("Model %s:\n", toupper(nm)))
  print(head(get(paste0("tuned_", nm))$all_results[, c("mean_rmse","max_depth",
        "eta","gamma","lambda","median_nrounds")], 3), row.names = FALSE)
}
```

---

# 10. Final Model Training and Evaluation

```{r final-models}
start_yr <- as.numeric(format(min(train_df$Date), "%Y"))
start_mo <- as.numeric(format(min(train_df$Date), "%m"))

fit_final <- function(train_data, test_data, feat_cols,
                      best_params, best_nrounds, label) {
  nrounds <- max(50L, best_nrounds)
  dtrain  <- xgb.DMatrix(data = as.matrix(train_data[, feat_cols]),
                          label = train_data$log_incidence)
  dtest   <- xgb.DMatrix(data = as.matrix(test_data[, feat_cols]))
  set.seed(42)
  suppressWarnings(
    model <- xgb.train(params = best_params, data = dtrain,
                       nrounds = nrounds, verbose = 0))
  pred_tr   <- bt(predict(model, dtrain), train_data$Population)
  pred_te   <- bt(predict(model, dtest),  test_data$Population)
  train_met <- calc_metrics(train_data$DengueCases, pred_tr)
  test_met  <- calc_metrics(test_data$DengueCases,  pred_te)
  cat(sprintf("%-40s nrounds=%d\n  Train: RMSE=%.0f MAE=%.0f R2=%.3f\n  Test:  RMSE=%.0f MAE=%.0f R2=%.3f\n",
              label, nrounds,
              train_met$rmse, train_met$mae, train_met$r2,
              test_met$rmse,  test_met$mae,  test_met$r2))
  list(model=model, feat_cols=feat_cols, train=train_met, test=test_met,
       pred_train=pred_tr, pred_test=pred_te, nrounds=nrounds)
}

# Model C: XGBoost + ARIMA residual correction
fit_final_hybrid <- function(train_data, test_data, feat_cols,
                              best_params, best_nrounds, label) {
  nrounds <- max(50L, best_nrounds)
  dtrain  <- xgb.DMatrix(data = as.matrix(train_data[, feat_cols]),
                          label = train_data$log_incidence)
  dtest   <- xgb.DMatrix(data = as.matrix(test_data[, feat_cols]))
  set.seed(42)
  suppressWarnings(
    model <- xgb.train(params = best_params, data = dtrain,
                       nrounds = nrounds, verbose = 0))

  # ARIMA on XGBoost training residuals
  resid_tr    <- ts(train_data$log_incidence - predict(model, dtrain),
                    start = c(start_yr, start_mo), frequency = 12)
  arima_resid <- auto.arima(resid_tr, seasonal = TRUE,
                             stepwise = FALSE, approximation = FALSE)
  resid_fc    <- forecast(arima_resid, h = nrow(test_data))
  cat(sprintf("  ARIMA residual model: (%s)\n",
              paste(arimaorder(arima_resid), collapse=",")))

  pred_log_tr <- predict(model, dtrain) + as.numeric(fitted(arima_resid))
  pred_log_te <- predict(model, dtest)  + as.numeric(resid_fc$mean)
  pred_tr     <- bt(pred_log_tr, train_data$Population)
  pred_te     <- bt(pred_log_te, test_data$Population)
  train_met   <- calc_metrics(train_data$DengueCases, pred_tr)
  test_met    <- calc_metrics(test_data$DengueCases,  pred_te)
  cat(sprintf("%-40s nrounds=%d\n  Train: RMSE=%.0f MAE=%.0f R2=%.3f\n  Test:  RMSE=%.0f MAE=%.0f R2=%.3f\n",
              label, nrounds,
              train_met$rmse, train_met$mae, train_met$r2,
              test_met$rmse,  test_met$mae,  test_met$r2))
  list(model=model, arima_resid=arima_resid, feat_cols=feat_cols,
       train=train_met, test=test_met,
       pred_train=pred_tr, pred_test=pred_te, nrounds=nrounds)
}

model_a <- fit_final(train_df, test_df, feat_a,
                     tuned_a$best_params, tuned_a$best_nrounds,
                     "Model A: AR only")
model_b <- fit_final(train_df, test_df, feat_b,
                     tuned_b$best_params, tuned_b$best_nrounds,
                     "Model B: AR + CBPFE")
model_c <- fit_final_hybrid(train_df, test_df, feat_c,
                            tuned_c$best_params, tuned_c$best_nrounds,
                            "Model C: XGB+ARIMA hybrid")
```

---

# 11. ARIMA-Family Baselines

```{r arima-baselines}
y_train_ts  <- ts(train_df$log_incidence,
                  start = c(start_yr, start_mo), frequency = 12)
exog_cols   <- c(season_cols, dlnm_effect_cols)
xreg_train  <- as.matrix(train_df[, exog_cols])
xreg_test   <- as.matrix(test_df[, exog_cols])
xreg_raw_tr <- as.matrix(train_df[, c("MeanTemp","Precipitation","Humidity","WindSpeed")])
xreg_raw_te <- as.matrix(test_df[,  c("MeanTemp","Precipitation","Humidity","WindSpeed")])

compute_vif <- function(xreg_mat) {
  if (is.null(xreg_mat) || ncol(xreg_mat) < 2) return(NULL)
  cor_mat  <- cor(xreg_mat, use = "complete.obs")
  vif_vals <- tryCatch(diag(solve(cor_mat)), error = function(e) NULL)
  if (!is.null(vif_vals)) names(vif_vals) <- colnames(xreg_mat)
  vif_vals
}

eval_arima <- function(fit, xreg_tr, xreg_te, label, train_data, test_data) {
  pred_tr  <- bt(as.numeric(fitted(fit)), train_data$Population)
  fc       <- if (is.null(xreg_te)) forecast(fit, h = nrow(test_data))
              else                   forecast(fit, xreg = xreg_te, h = nrow(test_data))
  pred_te  <- bt(as.numeric(fc$mean), test_data$Population)
  tr_m     <- calc_metrics(train_data$DengueCases, pred_tr)
  te_m     <- calc_metrics(test_data$DengueCases,  pred_te)
  lb       <- Box.test(residuals(fit), lag = 20, type = "Ljung-Box")
  cat(sprintf("%-45s  Train RMSE=%6.0f | Test RMSE=%6.0f R2=%.3f | LB p=%.3f\n",
              label, tr_m$rmse, te_m$rmse, te_m$r2, lb$p.value))

  vif_vals <- compute_vif(xreg_tr)
  max_vif  <- if (!is.null(vif_vals)) max(vif_vals) else NA_real_

  if (!is.na(max_vif)) {
    cat(sprintf("  VIF (max): %.2f %s\n", max_vif,
      ifelse(max_vif > 10, "[WARN: severe multicollinearity]",
      ifelse(max_vif > 5,  "[CAUTION: moderate]", "[OK]"))))
    vif_df <- data.frame(Variable = names(vif_vals), VIF = round(vif_vals, 2))
    print(head(vif_df[order(-vif_df$VIF), ], 5), row.names = FALSE)
  } else {
    cat("  VIF: N/A (no exogenous regressors)\n")
  }

  list(train = tr_m, test = te_m, pred_train = pred_tr, pred_test = pred_te,
       ljung_p = lb$p.value, fit = fit,
       max_vif = max_vif, vif_vals = vif_vals)
}

fit_arima        <- Arima(y_train_ts, order = c(2,1,2))
fit_sarima       <- Arima(y_train_ts, order = c(2,1,2),
                          seasonal = list(order=c(1,1,1), period=6))
fit_arimax       <- Arima(y_train_ts, order = c(2,1,2), xreg = xreg_train)
fit_sarimax      <- Arima(y_train_ts, order = c(2,1,2),
                          seasonal = list(order=c(1,1,1), period=6),
                          xreg = xreg_train)
fit_arimax_raw   <- Arima(y_train_ts, order = c(2,1,2), xreg = xreg_raw_tr)
fit_sarimax_raw  <- Arima(y_train_ts, order = c(2,1,2),
                          seasonal = list(order=c(1,1,1), period=6),
                          xreg = xreg_raw_tr)

res_arima   <- eval_arima(fit_arima,   NULL,        NULL,        "ARIMA(2,1,2)",            train_df, test_df)
res_sarima  <- eval_arima(fit_sarima,  NULL,        NULL,        "SARIMA(2,1,2)(1,1,1)[6]", train_df, test_df)
res_arimax  <- eval_arima(fit_arimax,  xreg_train,  xreg_test,   "ARIMAX(DLNM)",            train_df, test_df)
res_sarimax <- eval_arima(fit_sarimax, xreg_train,  xreg_test,   "SARIMAX(DLNM)",           train_df, test_df)
res_arimax_raw  <- eval_arima(fit_arimax_raw,  xreg_raw_tr, xreg_raw_te, "ARIMAX(weather)", train_df, test_df)
res_sarimax_raw <- eval_arima(fit_sarimax_raw, xreg_raw_tr, xreg_raw_te, "SARIMAX(weather)",train_df, test_df)
```

---

# 12. Results Tables

```{r results-tables}
# XGBoost model comparison
ml_comp <- data.frame(
  Model      = c("A: AR only", "B: AR + CBPFE", "C: XGB+ARIMA hybrid"),
  Features   = c(length(feat_a), length(feat_b), length(feat_c)),
  Train_RMSE = round(c(model_a$train$rmse, model_b$train$rmse, model_c$train$rmse)),
  Test_RMSE  = round(c(model_a$test$rmse,  model_b$test$rmse,  model_c$test$rmse)),
  Test_MAE   = round(c(model_a$test$mae,   model_b$test$mae,   model_c$test$mae)),
  Test_R2    = round(c(model_a$test$r2,    model_b$test$r2,    model_c$test$r2), 3),
  Ratio      = round(c(model_a$train$rmse/model_a$test$rmse,
                        model_b$train$rmse/model_b$test$rmse,
                        model_c$train$rmse/model_c$test$rmse), 2)
)
knitr::kable(ml_comp, caption = "Table 2: XGBoost model comparison (2024–2025 test set)")

# ARIMA comparison
arima_list  <- list(res_arima, res_sarima, res_arimax, res_sarimax,
                    res_arimax_raw, res_sarimax_raw)
arima_names <- c("ARIMA", "SARIMA", "ARIMAX (DLNM)", "SARIMAX (DLNM)",
                 "ARIMAX (weather)", "SARIMAX (weather)")
arima_comp  <- data.frame(
  Model      = arima_names,
  Train_RMSE = round(sapply(arima_list, function(r) r$train$rmse)),
  Test_RMSE  = round(sapply(arima_list, function(r) r$test$rmse)),
  Test_R2    = round(sapply(arima_list, function(r) r$test$r2), 3),
  LjungBox_p = round(sapply(arima_list, function(r) r$ljung_p), 3),
  Max_VIF    = round(sapply(arima_list, function(r) r$max_vif), 2)
)
knitr::kable(arima_comp, caption = "Table 3: ARIMA-family comparison (2024–2025 test set)")
```

---

# 13. Forecast Plots

```{r forecast-plots}
combined_df <- test_df %>%
  dplyr::select(Date, Actual = DengueCases) %>%
  mutate(
    ARIMA               = res_arima$pred_test,
    SARIMA              = res_sarima$pred_test,
    `ARIMAX (DLNM)`     = res_arimax$pred_test,
    `SARIMAX (DLNM)`    = res_sarimax$pred_test,
    `ARIMAX (weather)`  = res_arimax_raw$pred_test,
    `SARIMAX (weather)` = res_sarimax_raw$pred_test,
    `XGB A: AR only`    = model_a$pred_test,
    `XGB B: AR + CBPFE` = model_b$pred_test,
    `XGB C: Hybrid`     = model_c$pred_test
  ) %>%
  tidyr::pivot_longer(-Date, names_to = "Series", values_to = "Cases") %>%
  mutate(Series = factor(Series, levels = c(
    "Actual", "ARIMA", "SARIMA",
    "ARIMAX (DLNM)", "SARIMAX (DLNM)",
    "ARIMAX (weather)", "SARIMAX (weather)",
    "XGB A: AR only", "XGB B: AR + CBPFE", "XGB C: Hybrid")))

ggplot(combined_df, aes(x = Date, y = Cases, colour = Series, linetype = Series)) +
  geom_line(linewidth = 1.0) + geom_point(size = 1.2) +
  scale_colour_manual(values = c(
    "Actual"="black", "ARIMA"="#E69F00", "SARIMA"="#56B4E9",
    "ARIMAX (DLNM)"="#009E73", "SARIMAX (DLNM)"="#CC79A7",
    "ARIMAX (weather)"="#D55E00", "SARIMAX (weather)"="#0072B2",
    "XGB A: AR only"="#F0E442", "XGB B: AR + CBPFE"="#E31A1C",
    "XGB C: Hybrid"="#984EA3")) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 months") +
  labs(title = "Dengue Forecasts: 2024–2025 — All Models",
       subtitle = sprintf("Best ML: Model B (RMSE=%.0f) | Best ARIMA: SARIMAX DLNM (RMSE=%.0f)",
                          model_b$test$rmse, res_sarimax$test$rmse),
       x = "Date", y = "Dengue Cases") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(colour = guide_legend(nrow = 3), linetype = guide_legend(nrow = 3))
```

---

# 14. SHAP Explainability (Model B)

SHAP values are computed on the **test set feature matrix** to assess out-of-sample
feature contributions.

```{r shap-analysis}
x_test_b  <- as.matrix(test_df[, feat_b])
shap_b    <- shap.values(xgb_model = model_b$model, X_train = x_test_b)

cat("=== SHAP Feature Importance (Model B, Test Set) ===\n")
print(round(shap_b$mean_shap_score, 4))

shap_long_b <- shap.prep(shap_contrib = shap_b$shap_score, X_train = x_test_b)
shap.plot.summary(shap_long_b) +
  ggtitle("SHAP Summary — Model B: AR + CBPFE (Test Set 2024–2025)")

# Dependence plots for top 4 features
top4 <- names(shap_b$mean_shap_score)[seq_len(min(4, length(shap_b$mean_shap_score)))]
for (feat in top4) {
  print(shap.plot.dependence(shap_long_b, x = feat,
                              color_feature = "auto", smooth = TRUE, size0 = 1.5) +
        ggtitle(paste("SHAP Dependence:", feat)))
}
```

---

# 15. External Validation — Brazil

```{r external-validation-brazil}
# Load Brazil dataset (Coelho et al., 2024 — Zenodo DOI: 10.5281/zenodo.13328231)
brazil_raw          <- read.csv("dataset/brazil_monthly_dataset.csv")
brazil_raw$Date     <- as.Date(brazil_raw$month)
brazil_raw$Year     <- year(brazil_raw$Date)
brazil_raw$Month_num <- month(brazil_raw$Date)
names(brazil_raw)[names(brazil_raw) == "cases"]    <- "DengueCases"
names(brazil_raw)[names(brazil_raw) == "temp"]     <- "MeanTemp"
names(brazil_raw)[names(brazil_raw) == "rain"]     <- "Precipitation"
names(brazil_raw)[names(brazil_raw) == "humidity"] <- "Humidity"

# Annual population from World Bank
brazil_pop <- data.frame(
  Year       = 2010:2024,
  Population = c(193701929, 195284734, 196876111, 198478299, 200085127,
                 201675532, 203218114, 204703445, 206107261, 207455459,
                 208660842, 209550294, 210306415, 211140729, 211998573)
)
brazil_raw <- brazil_raw %>% left_join(brazil_pop, by = "Year")

# Z-score mapping: preserves relative variation, aligns to Bangladesh feature space
scale_to_bd <- function(bz_vals, bd_vals)
  (bz_vals - mean(bz_vals, na.rm=TRUE)) / sd(bz_vals, na.rm=TRUE) *
  sd(bd_vals, na.rm=TRUE) + mean(bd_vals, na.rm=TRUE)

brazil <- brazil_raw
brazil$MeanTemp      <- scale_to_bd(brazil$MeanTemp,      train_df$MeanTemp)
brazil$Precipitation <- scale_to_bd(brazil$Precipitation, train_df$Precipitation)
brazil$Humidity      <- scale_to_bd(brazil$Humidity,      train_df$Humidity)
brazil$WindSpeed     <- scale_to_bd(brazil$WindSpeed,     train_df$WindSpeed)

# Rebuild crossbases with identical spline specs
brazil           <- as.data.frame(brazil)
brazil$Month_index <- seq_len(nrow(brazil))
cb_MT_bz <- crossbasis(brazil$MeanTemp,      lag=lag_period,
  argvar=list(fun="ns",df=2,Boundary.knots=pred_MeanTemp_range),
  arglag=list(fun="ns",df=2))
cb_PR_bz <- crossbasis(brazil$Precipitation, lag=lag_period,
  argvar=list(fun="ns",df=3,Boundary.knots=pred_Precip_range),
  arglag=list(fun="ns",df=3))
cb_HU_bz <- crossbasis(brazil$Humidity,      lag=lag_period,
  argvar=list(fun="ns",df=3,Boundary.knots=pred_Humidity_range),
  arglag=list(fun="ns",df=3))
cb_WS_bz <- crossbasis(brazil$WindSpeed,     lag=lag_period,
  argvar=list(fun="ns",df=2,Boundary.knots=range(df$WindSpeed)),
  arglag=list(fun="ns",df=2))

bz_rows     <- (lag_period + 1):nrow(brazil)
brazil_fit  <- brazil[bz_rows, ]
cb_MT_bz    <- subset_cb(cb_MT_bz, bz_rows)
cb_PR_bz    <- subset_cb(cb_PR_bz, bz_rows)
cb_HU_bz    <- subset_cb(cb_HU_bz, bz_rows)
cb_WS_bz    <- subset_cb(cb_WS_bz, bz_rows)

# Extract partial effects using Bangladesh-trained coefficients (Step 3)
extract_effect <- function(dlnm_obj, cb_bz) {
  cf       <- coef(dlnm_obj$model)
  cb_coefs <- cf[grep("^cb_oos", names(cf))]
  stopifnot(ncol(cb_bz) == length(cb_coefs))
  as.numeric(cb_bz %*% cb_coefs)
}
brazil_fit$dlnm_effect_MT <- extract_effect(dlnm_MT, cb_MT_bz)
brazil_fit$dlnm_effect_PR <- extract_effect(dlnm_PR, cb_PR_bz)
brazil_fit$dlnm_effect_HU <- extract_effect(dlnm_HU, cb_HU_bz)
brazil_fit$dlnm_effect_WS <- extract_effect(dlnm_WS, cb_WS_bz)

brazil_fit$log_incidence <- log1p(brazil_fit$DengueCases / brazil_fit$Population * 1e5)
for (k in c(1:6, 12))
  brazil_fit[[paste0("lag", k, "_dengue")]] <- dplyr::lag(
    log1p(brazil_fit$DengueCases), k)
brazil_fit$sin_month <- sin(2 * pi * brazil_fit$Month_num / 12)
brazil_fit$cos_month <- cos(2 * pi * brazil_fit$Month_num / 12)

brazil_valid <- brazil_fit %>%
  tidyr::drop_na(all_of(c(ar_cols, dlnm_effect_cols, "log_incidence")))
cat("Brazil rows for prediction:", nrow(brazil_valid), "\n")

pred_log_bz   <- predict(model_b$model,
                          xgb.DMatrix(as.matrix(brazil_valid[, feat_b])))
pred_cases_bz <- bt(pred_log_bz, brazil_valid$Population)

bz_m <- calc_metrics(brazil_valid$DengueCases, pred_cases_bz)
cat(sprintf("\nBrazil external validation:\n  RMSE=%.0f | MAE=%.0f | R2=%.3f\n",
            bz_m$rmse, bz_m$mae, bz_m$r2))
cat(sprintf("  Spearman r = %.3f | Pearson r = %.3f\n",
    cor(brazil_valid$DengueCases, pred_cases_bz, method="spearman"),
    cor(brazil_valid$DengueCases, pred_cases_bz, method="pearson")))

# Time-series plot (log scale)
bz_plot <- brazil_valid %>%
  dplyr::select(Date, Actual = DengueCases) %>%
  mutate(Predicted = pred_cases_bz) %>%
  tidyr::pivot_longer(-Date, names_to="Series", values_to="Cases")

ggplot(bz_plot, aes(x=Date, y=Cases, colour=Series, linetype=Series)) +
  geom_line(linewidth=1.0) + geom_point(size=1.2) +
  scale_colour_manual(values=c("Actual"="black","Predicted"="#E31A1C")) +
  scale_y_log10(labels=scales::comma) +
  scale_x_date(date_labels="%b %Y", date_breaks="12 months") +
  labs(title="External Validation: Model B on Brazil (log scale)",
       subtitle=sprintf("Spearman r=%.3f — epidemic timing detection",
                        cor(brazil_valid$DengueCases, pred_cases_bz, method="spearman")),
       x="Date", y="Dengue Cases (log scale)") +
  theme_minimal(base_size=12) +
  theme(legend.position="bottom", plot.title=element_text(face="bold"),
        axis.text.x=element_text(angle=45, hjust=1))
```

---

# Session Info

```{r session-info}
sessionInfo()
```
